---
title: "HW2 ggplot"
subtitle: BIOL 607
format: html
---

## 1. Setup

```{r}
#| warning: false 
library(dplyr)
library(readr)
library(ggplot2)
library(forcats)
library(RColorBrewer)
library(ggthemes)
pacman::p_load_gh("Mikata-Project/ggthemr")

theme_set(theme_bw(base_size=12))

ice <- read_csv("http://biol607.github.io/homework/data/NH_seaice_extent_monthly_1978_2016.csv") |>
  mutate(Month_Name = factor(Month_Name),
         Month_Name = fct_reorder(Month_Name, Month))

str(ice)
```

## 2. Boxplots
```{r}
ggthemr('light')

boxp <- ggplot(data = ice,
               mapping = aes(x = Month_Name,
                             y = Extent,
                             fill = Month))

finalbp <- boxp + 
  geom_boxplot(notch = TRUE, notchwidth = 0.8, 
               outlier.color = "red", outlier.fill = "black", outlier.size = 1) +
  theme(legend.position = "none") +
  scale_fill_viridis_c(option = "H") +
  labs(title = "Variability in Sea Ice per Month",
       x = "Month")

finalbp
```

## 3. Ridgeline (ggridges)
```{r}
#| message: false
#| warning: false
library(ggridges)

ridge <- ggplot(ice, aes(x = Extent, y = Month_Name, fill = ..x..)) +
  geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01) +
   scale_fill_viridis_c(direction = -1, option = "A") +
  theme(legend.position = "None") +
  labs(Title = "Variability in Sea Ice per Month",
       y = "Month")

ridge
```

## 4. Groups from the Continuous
```{r}
seasons <- ggplot(ice, aes(y = Year, x = Extent, color = Extent, group = Month)) +
  geom_line() +
  scale_color_viridis_c(direction = -1, option = "B") +
  facet_wrap(vars(cut_interval(Month, n=4)))
seasons +
  theme(legend.position = "None") +
  labs(title = "Variability in Sea Ice \nThroughout the Years", 
       caption = "Organized by Quarters of the Year")
```

## 5. Using Colors

#### a. ColorFindr
```{r}
library(colorfindr)

aab <- get_colors("https://upload.wikimedia.org/wikipedia/commons/8/89/Flag_of_Antigua_and_Barbuda.svg") |>
  make_palette(n = 5, show = TRUE)
```
Antigua and Barbados Flag Color Palette created by colorfindr

#### b. Line Plot
```{r}
library(ggiraph)

colors <- ggplot(ice, aes(x = Month_Name, y = Extent, group = Year, color = Year)) +
  geom_line() +
  scale_color_gradientn(colors = aab) +
  geom_point_interactive(aes(x = Month, y = Extent, tooltip = Year, color = Year, data_id = Month_Name), size = 1) +
  theme_get() +
  theme(legend.position = "None") +
  labs(title = "Variability in Sea Ice",
       x = "Month", 
       caption = "Interact with Points for Exact Year")


x <- girafe(ggobj = colors)
if (interactive()) {print(x)}
x
```


## gganimate
```{r}
#| message: false
#| warning: false
library(gganimate)

p1 <- ggplot(ice, aes(x = Month_Name, y = Extent, group = Year, color = Year))+
  geom_line() +
  labs(title = 'Variability in Sea Ice in Year: {frame_time}', x = 'Month', y = 'Extent') +
  transition_time(Year) +
  scale_color_gradientn(colors = aab) + 
  theme_gray()

animate(p1)
```

## Impress Myself
```{r}
impr <- ggplot(ice, aes(x = Extent, y = Month_Name, fill = Month)) + 
  geom_bar(stat = "identity") +
  transition_time(Year) +
  labs(title = 'Variability in Sea Ice in Year: {frame_time}', x = 'Extent', y = 'Month') +
   shadow_trail(alpha = 0.2, shape = 4, max_frames = 100) +
  scale_fill_distiller(palette = "Set2") +
  theme_solarized(light = FALSE) +
  theme(legend.position="none")

animate(impr)
```

## TidyTuesday - CDC DataSets
#### 2025 Week 6 / 2-11-2025

```{r}
#| message: false
#| warning: false
library(readr)
library(dplyr) 
library(tidyr)

mal <- read_csv("data/Number_of_Reported_Malaria_Cases_by_County_US_2016.csv")

MATX <- filter(mal, FIPS %in% c("6085","9003", "25025", "48113", "25027", "25017", "48201", "48135"))
CACT <- filter(mal, COUNTY %in% c("SANTA CLARA", "HARTFORD", "Dallas", "Suffok", "ALAMEDA","SACRAMENTO") & STATE %in% c("CALIFORNIA", "CONNECTICUT", "TEXAS", "MASSACHUSETTS"))

compar <- rbind(MATX, CACT)
```
Using Centers for Disease Control and Prevention's (CDC) data set on Malaria Outbreak in 2016 to visualize what State/Region would impact Malaria susceptibility (In States I have lived/wish to). 

```{r}
compar_mut <- mutate(compar, COUNTY = fct_reorder(COUNTY, MAL_FREQ_2016))

compar_gg <- ggplot(compar_mut, mapping = aes(x=STATE, y=MAL_FREQ_2016, fill = COUNTY, tooltip = COUNTY, data_id = COUNTY)) +
  geom_bar_interactive(stat = "identity", position = "dodge") +
      theme_dark() +
  labs(Title = "Malaria Frequency 2016 with Dantony's Counties of Interest", 
       x = "State",
       y = "Malaria Frequency",
       caption = "Dantony has Lived, Visited, or Interested in these Counties for Residence") +
  scale_fill_brewer(palette = "RdYlGn", direction = -1) +
  ggtitle("Malaria Frequency 2016 in Dantony's Areas of Interest") 
  
 


compar_int <- girafe(ggobj = compar_gg)
if (interactive()) {print(compar_int)}
compar_int
```

## Meta Questions
1) Within this week, I have thought about multiple different ways to create visualizations within the same data. This thinking process does not occur often for me but through repetition and using helpful chart selections I am able to continue expanding my knowledge on the effectiveness of data visualization. Within ggplot I was not aware of the extent of customization that is offered through ggplot, r, and other packages even though I knew the basics of ggplot. 

2) My favorite aspect of data visualization is the creative freedom that one has regardless of duplicate data. This allows for a lot of fun aspects to be added within the visualization, truly making a visualizer that is easier to visualize. 

3) It took me roughly 5 hours throughout the week to complete this assignment. I do not believe this was due to how the homework was assigned but due to my curiousity causing me to explore new things that may be too advance for me. 

4) I would give my self a studdicient leaning toward strong assessment on this assignment. This assessment is due to attacking each question to it's fullest even though it may have taken a lot of time to figure out what I would exactly like. I also attemped everything present with full completion.