---
title: "HW3 Tidy Data"
subtitle: "BIOL 607"
format: html
---

# Tidy Data 

### 1 - Load
```{r}
#| message: false

library(readr)

kelp_nd <- read_csv("https://github.com/kelpecosystems/observational_data/blob/master/cleaned_data/keen_cover.csv?raw=true")

kelp_dwn <- read_csv("data/keen_cover.csv")
```

## 2 - Format
##### This data is in long format, due to each variable is only listed once thus causing  more rows of data making it longer. Usually the variables are combine with different    variables such as time which extends the columns thus making it a wide format.\

## 3 - Check it Out !
### 3a. Amount of Sites per PI
```{r}
#| message: false
library(dplyr)
library(tidyr)
kelp_pi <- 
  kelp_nd |>
  select(PI, SITE)
```
##### Byrnes
```{r}
#Code Repeated for other PI's
kelp_byrnes <-
  kelp_pi |>
  filter(PI == "Byrnes") |>
  distinct(SITE, .keep_all = TRUE)
n_distinct(kelp_byrnes$SITE)
```
##### Dijkstra
```{r}
#| echo: false

kelp_dijkstra <- 
  kelp_pi |>
  filter(PI == "Dijkstra")
n_distinct(kelp_dijkstra$SITE)
```

##### Grabowski
```{r}
#| echo: false

kelp_grabowski <- 
  kelp_pi |>
  filter(PI == "Grabowski")
n_distinct(kelp_grabowski$SITE)
```

##### Humphries
```{r}
#| echo: false

kelp_humphries <- 
  kelp_pi |>
  filter(PI == "Humphries") |>
  distinct(SITE, .keep_all = TRUE)
n_distinct(kelp_humphries$SITE)
```

##### Pemaquid
```{r}
#| echo: false

kelp_pemaquid <- 
  kelp_pi |>
  filter(PI == "Pemaquid")
n_distinct(kelp_pemaquid$SITE)
```

##### Hurricane
```{r}
#| echo: false

kelp_hurricane <- 
  kelp_pi |>
  filter(PI == "Hurricane")
n_distinct(kelp_hurricane$SITE)
```

### 3b. Years of Data per Site
```{r}
#Code Repeated for other Sites
kelp_site <- 
  kelp_nd |>
  select(SITE, YEAR)
n_distinct(kelp_site$SITE)
```

##### Baker North
```{r}
#| echo: false

kelp_bakern <- 
  kelp_site |>
  filter(SITE == "Baker North") |>
  arrange(desc(YEAR)) |>
  distinct(YEAR, .keep_all = TRUE)
kelp_bakern
```

##### Baker South
```{r}
#| echo: false

kelp_bakers <- 
  kelp_site |>
  filter(SITE == "Baker South") |>
  arrange(desc(YEAR)) |>
  distinct(YEAR, .keep_all = TRUE)
kelp_bakers
```

##### Calf Island
```{r}
#| echo: false

kelp_calf_island <- 
  kelp_site |>
  filter(SITE == "Calf Island") |>
  arrange(desc(YEAR)) |>
  distinct(YEAR, .keep_all = TRUE)
kelp_calf_island
```

##### Little Brewster
```{r}
#| echo: false

kelp_little_brew <- 
  kelp_site |>
  filter(SITE == "Little Brewster") |>
  arrange(desc(YEAR)) |>
  distinct(YEAR, .keep_all = TRUE)
kelp_little_brew
```

##### NE Appledore
```{r}
#| echo: false

kelp_ne_appledore <- 
  kelp_site |>
  filter(SITE == "NE Appledore") |>
  arrange(desc(YEAR)) |>
  distinct(YEAR, .keep_all = TRUE)
kelp_ne_appledore
```

##### NW Appledore
```{r}
#| echo: false

kelp_nw_appledore <- 
  kelp_site |>
  filter(SITE == "NW Appledore") |>
  arrange(desc(YEAR)) |>
  distinct(YEAR, .keep_all = TRUE)
kelp_nw_appledore
```

##### SW Appledore
```{r}
#| echo: false

kelp_sw_appledore <- 
  kelp_site |>
  filter(SITE == "SW Appledore") |>
  arrange(desc(YEAR)) |>
  distinct(YEAR, .keep_all = TRUE)
kelp_sw_appledore
```

##### Nubble Lighthouse
```{r}
#| echo: false

kelp_nubble_lh <- 
  kelp_site |>
  filter(SITE == "Nubble Lighthouse") |>
  arrange(desc(YEAR)) |>
  distinct(YEAR, .keep_all = TRUE)
kelp_nubble_lh
```

##### Nahant
```{r}
#| echo: false

kelp_nahant <- 
  kelp_site |>
  filter(SITE == "Nahant") |>
  arrange(desc(YEAR)) |>
  distinct(YEAR, .keep_all = TRUE)
kelp_nahant
```

##### Fort Weatherill
```{r}
#| echo: false

kelp_ft_weatherill <- 
  kelp_site |>
  filter(SITE == "Fort Weatherill") |>
  arrange(desc(YEAR)) |>
  distinct(YEAR, .keep_all = TRUE)
kelp_ft_weatherill
```

##### King's Beach
```{r}
#| echo: false

kelp_kings_beach <- 
  kelp_site |>
  filter(SITE == "King's Beach") |>
  arrange(desc(YEAR)) |>
  distinct(YEAR, .keep_all = TRUE)
kelp_kings_beach
```

##### Hurricane Island
```{r}
#| echo: false

kelp_hurricane_island <- 
  kelp_site |>
  filter(SITE == "Hurricane Island") |>
  arrange(desc(YEAR)) |>
  distinct(YEAR, .keep_all = TRUE)
kelp_hurricane_island
```

##### Permaquid
```{r}
#| echo: false
#|message: false
kelp_pemaquid_yr <- 
  kelp_site |>
  filter(SITE == "Pemaquid") |>
  arrange(desc(YEAR)) |>
  distinct(YEAR, .keep_all = TRUE)
kelp_pemaquid_yr
```

### 3c. Impress Yourself / Figure
```{r}
library(ggplot2)
site_year <- 
  kelp_site |>
  ggplot(mapping = aes(x = YEAR, y = SITE, fill = SITE)) +
  geom_tile() +
  labs(title = "Years Site in Use",
       y = "Site",
       x = "Year") +
  theme_linedraw() +
  theme(legend.position = "None") +
  scale_x_continuous(breaks=seq(2014,2022,by=1))

site_year
```

## 4 Kelp
### 4a. 
```{r}
kelp_trim <- 
  kelp_nd |>
  select(YEAR, SITE, TRANSECT, PERCENT_COVER, FAMILY, SPECIES)
kelp_trim
```

### 4b. 
```{r}
kelp_trimer <- 
  kelp_trim |>
  filter(FAMILY == "Laminariaceae") |>
  select(!FAMILY)
kelp_trimer
```

### 4c. 
##### No there are two measurements per species transect each year.

### 4d.
```{r}
#| message: false
#| warning: false

kelp_cover <- 
kelp_trimer |>
  group_by(TRANSECT,SITE,YEAR) |>
  summarize(sum_cover = sum(as.numeric(PERCENT_COVER)))
kelp_cover
```
### 4e.
```{r}
#| message: false
#| warning: false

kelp_plot <- 
  kelp_trimer|>
  ggplot(mapping = aes(x = YEAR, y = PERCENT_COVER, color = SITE)) +
  facet_wrap(~SPECIES) +
  stat_summary(geom = "line") +
  theme_dark() +
  scale_color_manual(values = c("firebrick4","Orange", "Yellow", "seagreen", "Navy", "Purple", "White", "plum1", "skyblue", "yellowgreen", "lightcoral", "deeppink3","burlywood", "aquamarine2")) +
  labs(title = "Timeseries of Kelp by Site",
       y = "Percent Cover",
       x = "Year") 
kelp_plot
```
##### I noticed that the *Laminaria digitata* species has a lower percent cover as well as less data collected, compared to *Saccharina latissima* where the percent cover range is larger and more data available.

## 5 Wide Relationships
### 5a. 
```{r}
#| message: false
#| warning: false

kelp_wide_prep <- 
  kelp_trimer |>
  group_by(TRANSECT,SITE,YEAR, SPECIES) |>
  summarize(sum_cover = sum(PERCENT_COVER, na.rm = TRUE))

kelp_wide <- 
  kelp_wide_prep |>
  group_by(SPECIES, sum_cover) |>
  pivot_wider(names_from = SPECIES,
              values_from = sum_cover)
kelp_wide[is.na(kelp_wide)] = 0
kelp_wide
```
### 5c. 
```{r}
#| message: false
#| warning: false
#| 
library(ggtext)
library(ggthemes)

sp_compar <- 
  kelp_wide |>
  ggplot() +
  geom_smooth(aes(x = YEAR, y = `Saccharina latissima`, color = "skyblue", method = "lm")) +
  geom_smooth(aes(x = YEAR, y = `Laminaria digitata`, color = "lightpink", method = "lm")) +
  labs(title = "Percent Cover Relationship between Species",
       subtitle = "<span style = 'color: skyblue;'>Blue</span> = Saccharina latissima    <span style = 'color: pink;'>Pink</span> = Laminaria digitata",
       x = "Year",
       y = "Percent Cover") +
  theme_stata() +
  theme(plot.subtitle = element_markdown(),
        legend.position = "None")
sp_compar
```

### 5d. 
```{r}
#| message: false
#| warning: false

kelp_wide_prep[is.na(kelp_wide_prep)] = 0

long_graph <- 
  kelp_wide_prep|>
  ggplot(mapping = aes(x = YEAR, y = sum_cover, color = SITE)) +
  facet_wrap(~SPECIES) +
  stat_summary(geom = "line") +
  theme_dark() +
  scale_color_manual(values = c("firebrick4","Orange", "Yellow", "seagreen", "Navy", "Purple", "White", "plum1", "skyblue", "yellowgreen", "lightcoral", "deeppink3","burlywood", "aquamarine2")) +
  labs(title = "Timeseries of Kelp by Site") 
long_graph
```
##### I noticed in this there are slight changes to the lines present, especially within the *Saccharina latissima* species that seems to reach a high sum cover with the NA at zero. The zeros do not completely change the story but it does have an affect on the way we anaylize it. 

## Meta Questions

### 1)
##### Working with data found in the wild is sonmething that I have done in the past but within the lectures I was able to gain a better understanding on manipulating data. When I first examined the data, I noticed the sheer amount of data present was daunting but as you break it down it become a lot more comprehensible.

### 2)
##### As stated before this is something that I have done previous but wihtin week 3 I learned a lot more detail and now feel like I have an understanding of data manipulation just need more practice.

### 3)
##### In the past I have always created and work with long form datasets and this was truly my first introduction to wide datasets. From the Browman and Woo paper I learned a lot more about proper data organization such as using good names, one value in cell with no empty, backups, etc.

### 4)
##### This assignment took a little longer than expected, taking about 3 hours. This may due to issues I had with plotting as well as the repition.

### 5)
I would give myself a sufficient on this assignment, as I did not go beyond on most of the questions as well I struggled at times. 
 