---
title: "HW1 Intro to R"
subtitle: BIOL 607
format: html
---

# Earthquakes off Fiji

### 1 - str() & summary()

With the **str()** function you are able to view the amount of observations in data frame, view the amount fo variables/row, and view the class of certain objects. **summary()** can be used to find different aggregate functions of each variable given in data frame.

```{r}
data(quakes)
str(quakes)
summary(quakes)
```

### 2 - Showing Entirety of Column

```{r}
quakes$long
```

### 3 - Finding Unique Values

```{r}
unique(quakes$long)
length(unique(quakes$long))
```

### 4 - Range

```{r}
range(quakes$depth)
```

### Impressing Myself

```{r}
max(quakes$mag)
quakes[quakes$mag == 6.4,]
```

The largest magnitude was recorded at station 122 on row 152

# LaTeX Syntax

### 5 - LaTex Mathmatical Symbols

$$
x = \mu
$$

### 6 - Difference of Equation Presenting

The difference between the two is, \$ stays inline compared to \$\$ the equation is displayed $Y_i = 3X_i$ $$
Y_i = 3X_i
$$

### 7 - Superscripts

$$
a^2 = b^2 + c^2
$$

### Impress Myself

$$
y_i \sim \mathcal{N}(\hat{y}_i,\sigma^2) 
$$


## Meta Questions

1. This was not my first experience with R as I have used it slightly within a Evolution course, but throughout the introduction to R week I was slowly regain memories of how to properly use R. All though it was not the most challenging problems, I was introduced to new concepts and fleshed out the basics in order for me to gain a true comprehension of R. 

2. I enjoy having the clear opportunity to learn more outside of class especially when it is more self-sufficient just like the LaTeX portion of the assignment. The LaTex portion was not too difficult, although I did struggle to find certain LaTex notations as I was looking for more specific characters. I also do see the future applicatin within LaTeX, especially when it comes to printing mathematical equations. 

3. It took me roughly an hour to fully complete the assignment, as well as about an hour or so to fully set up my GitHub.

4. I would give myself a strong assessment on this assignment, as I was able to complete each portion to its fullest while fully comprehending what process were taking place.  